Optimizing the power of adaptive seamless designs using Bayesian Optimization

### 2020-12-09 main.tex
reading 3

Bayesian Optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
Then it becomes infeasible to evaluate the black-box for thousands of configurations.

\subsection{General principle}
BO solves an optimization problem within a bounded search space $\Theta$:
\[
\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta),
\]
where $f(\theta)$ denotes the evaluation of the black-box with the input configuration $\theta$.
To reduce the number of evaluations on $f$ the key idea of BO is to only evaluate values of $\theta$ that are estimated to lead to a small value of $f(\theta)$.
The estimate is generated by a so called \emph{surrogate model}.
Typically, this is a regression model that predicts the outcome of $f$ based on previous evaluations of $f$.
First, an initial design of already evaluated configurations is needed.
Then, iteratively, the BO algorithm fits the surrogate on the previous evaluations, proposes a new configuration $\theta$ and evaluates it on $f$.
The steps are repeated until a budget is exhausted.

The proposal of a new configuration $\theta$ is obtained by maximizing a so called acquisition function.
The acquisition function guides the search to promising new regions in the search space $\Theta$ by facilitating the mean prediction $\hat{\mu}(\theta)$ as well as the uncertainty prediction $\hat{s}^2(\theta)$ of the surrogate.
It balances between exploration of not yet evaluated regions in $\Theta$ and exploitation.
The acquisition function achieves exploration by assigning high values to areas where the surrogate predicts a high uncertainty.
Exploitation is achieved by assigning high values to areas of $\theta$ where the surrogate predicts a high outcome of $f$.
For deterministic functions the expected improvement~\citep{jones_efficient_1998} is arguably the most popular acquisition function.

\subsection{Acquisition function for noisy outcomes}
Since this work focuses on optimization of a stochastic simulation, we assume that our non-deterministic optimization problem can be formulated as follows:
\begin{equation}
\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon, \, \text{ with } \, \varepsilon \sim N(0, \sigma^2_{n}) %\text{ i.i.d.}
\end{equation}
In this setting, the expected improvement is not feasible, as shown in \citet{huang_global_2006}.
They propose to use the augmented expected improvement instead.
Therefore, in a first step, we calculate the \emph{effective best solution}:
\begin{equation}
  \theta^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta) - c \cdot \hat{s}(\theta),
\end{equation}
with $c$ as a tuning parameter that is usually set to 1 and $\mathcal{D}$ as the design that contains all previously evaluated values of $\theta$.
The effective best point is the pessimistic estimate of the best observed outcome so far.
In the final step, we calculate the augmented expected improvement~\citep{huang_global_2006} as follows:
\begin{multline}
  \label{eq:aei}
  \operatorname{AEI}(\theta) = \left( \hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast}) \right) \cdot  \Phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{\hat{s}(\theta)} \biggr) \\
   + \hat{s}(\theta) \cdot \phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{\hat{s}(\theta)} \biggr) \cdot \biggl(1 - \underbrace{\frac{\sigma_{n}}{\sqrt{\sigma_{n}^2 + \hat{s}^2(\theta)}}}_{\text{correction}}\biggr),
\end{multline}
where $\Phi$ and $\phi$ are the distribution and density function of the standard normal distribution, and $\sigma_{n}^2$ denotes the random error (nugget effect) in the Kriging model.
The formula is derived under the assumption, that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
This assumption is also reflected by using a Gaussian process regression as a surrogate.
The intuition in formula (\ref{eq:aei}) is that the first summand favors exploitation while the second summand favors exploration.
The correction term is necessary to avoid exploration in areas where the model uncertainty $\hat{s}^2(\theta)$ equals the random error $\sigma_{n}^2$, because in this case further evaluations will not decrease the model uncertainty.

\subsection{Selection of the best point from noisy outcomes}
\label{ssec:best_point}
If we chose the configuration $\theta$ from all previously evaluated configurations according to the best outcome of $f$ as the optimization result, we would be overly optimistic.
It is likely that a single best outcome can be partially attributed to the random error $\sigma_{n}^2$.
Instead, we are interested in the optimum of the true posterior mean.
Therefore, we employ the surrogate estimate of the posterior mean for each evaluated configuration to cancel out the noise.
The configuration for which the surrogate estimates the best outcome is then returned as the optimization result $\hat{\theta}^\ast$:
\begin{equation}
  \hat{\theta}^{\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta),
\end{equation}
%To obtain a final outcome $\nu^\ast = f(\theta^\ast)$ that is independent of the optimization process, $f(\theta^\ast)$ is calculated again.
Using the stochastic outcome $\hat{\nu}^\ast := f(\hat{\theta}^\ast) $ observed during the optimization process would still potentially lead to an overly optimistic result.
Therefore an independent calculation of $f(\hat{\theta}^\ast)$ should be conducted to obtain a fair estimate of the true value of the best outcome.

@@

To compare \emph{BO} with a grid search that uses approximately the same time (see~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{BO} (see~\ref{tab:table_time}), with only a small advantage for \emph{Grid Small}.
Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{BO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
In the first case, the absolute differences are negligible as all methods lead to a high power.
For the second case we will later show, that the optimal configuration is close to the border of the search space.
The higher resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to the stochasticity.
These points are not included in the coarse grid.
In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.
The advantage of \emph{BO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect:paper2}, $n_{\text{treat}} = 500$.

@@

Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that each single optimization run leads to a results closer to the mean result.
Furthermore, the average optimization curve indicates that less iterations can be sufficient to obtain an outcome that is comparable to \emph{Grid},
It has to be noted, that this plot reports the best observed outcome so far and not the one that is chosen as the final result.
So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{BO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{BO} for the low and the high value of $n_\text{sim}$.
\begin{figure}[htb]
\centering
  \begin{minipage}{0.62\textwidth}
    \centering
    \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
    \captionof{figure}{Optimization progress for \emph{BO} in each stochastic repetition in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
    \label{fig:plot_opt_path_5000}
  \end{minipage}\hspace{0.02\textwidth}%
  \begin{minipage}{0.35\textwidth}
    \centering
    \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y_5000.pdf}%
    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_{\text{treat}} = 2000$ with 100 stochastic repetitions and different values for \texttt{nsim}}
    \label{fig:plot_boxplot_valid_y_5000}
  \end{minipage}
\end{figure}
For all methods the IQR displayed in the Box plots is much smaller for $n_\text{sim} = 5000$, indicating that the found solutions vary less within the 100 stochastic replications.
Also the absolute differences between the methods seem even more negligible.
This demonstrates, that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
But as an increased value of $n_\text{sim}$ also increases the runtime linearly, increasing $n_\text{sim}$ is not always a feasible choice if the goal is to find the optimal configuration quickly.
We conclude with the observation, that \emph{BO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulation.

@@

% subsection Summary
When planning clinical trials, determining patient group sizes and test methods is a crucial task.
The main goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
Either because large computational resources are required for an exhaustive exploration of all different configurations, or because it involves a tedious manual manual try and error process.
In this work we applied Model-based Optimization to quickly find a clinical trial setup with high power from a large number of setup options.
We compared out approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
BO was able to find competitive setups in a fraction of the time.

>>>>

Bayesian Optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
As then it becomes infeasible to evaluate the black-box for thousands of configurations.

\subsection{General principle}
BO solves an optimization problem within a bounded search space $\Theta$:
\[
\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta),
\]
where $f(\theta)$ denotes the evaluation of the black-box with the input configuration $\theta$.
To reduce the number of evaluations on $f$ the key idea of BO is to only evaluate values of $\theta$ that are estimated to lead to a high value of $f(\theta)$.
The estimate is generated by a so called \emph{surrogate model}.
Typically, this is a regression model that predicts the outcome of $f$ based on previous evaluations of $f$.
First, an initial design of already evaluated configurations is needed.
Then, iteratively, the BO algorithm fits the surrogate on the previous evaluations, proposes a new configuration $\theta$ and evaluates it on $f$.
The steps are repeated until a budget is exhausted.

The proposal of a new configuration $\theta$ is obtained by maximizing a so called acquisition function.
The acquisition function guides the search to promising new regions in the search space $\Theta$ by facilitating the mean prediction $\hat{\mu}(\theta)$ as well as the uncertainty quantification $s^2(\theta)$ of the surrogate.
It balances between exploration of not yet evaluated regions in $\Theta$ and exploitation.
The acquisition function achieves exploration by assigning high values to areas where the surrogate predicts a high uncertainty.
Exploitation is achieved by assigning high values to areas of $\theta$ where the surrogate predicts a high outcome of $f$.
For deterministic functions the expected improvement~\citep{jones_efficient_1998} is arguably the most popular acquisition function.

\subsection{Acquisition function for noisy outcomes}
Since this work focuses on optimization of a stochastic simulation, we assume that our non-deterministic optimization problem can be formulated as follows:
\begin{equation}
\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta) + \varepsilon, \, \text{ with } \, \varepsilon \sim N(0, \sigma^2) %\text{ i.i.d.}
\end{equation}
In this setting, the expected improvement is not feasible, as shown in \citet{huang_global_2006}.
They propose to use the augmented expected improvement instead.
Therefore, in a first step, we calculate the \emph{effective best solution}:
\begin{equation}
  \theta^{\ast\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta) - c \cdot s(\theta),
\end{equation}
with $c$ as a tuning parameter that is usually set to 1 and $\mathcal{D}$ as the design that contains all previously evaluated values of $\theta$.
The effective best point is the pessimistic estimate of the best observed outcome so far.
In the final step, we calculate the augmented expected improvement~\citep{huang_global_2006} as follows:
\begin{multline}
  \label{eq:aei}
  \operatorname{AEI}(\theta) = \left( \hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast}) \right) \cdot  \Phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{s(\theta)} \biggr) \\
   + s(\theta) \cdot \phi \biggl( \frac{\hat{\mu}(\theta) - \hat{\mu}(\theta^{\ast\ast})}{s(\theta)} \biggr) \cdot \biggl(1 - \underbrace{\frac{\sigma}{\sqrt{\sigma^2 + s^2(\theta)}}}_{\text{correction}}\biggr),
\end{multline}
where $\Phi$ and $\phi$ are the distribution and density function of the standard normal distribution, and $\sigma^2$ denotes the random error (nugget effect) in the Kriging model.
The formula is derived under the assumption, that at each point $\theta \in \Theta$ the posterior of the function outcome follows a normal distribution.
This assumption is also reflected by using a Gaussian process regression as a surrogate.
The intuition in formula (\ref{eq:aei}) is that the first summand favors exploitation while the second summand favors exploration.
The correction term is necessary to avoid exploration in areas where the model uncertainty $s^2(\theta)$ equals the random error $\sigma^2$, because in this case further evaluations will not decrease the model uncertainty.

\subsection{Selection of the best point from noisy outcomes}
\label{ssec:best_point}
If we chose the configuration $\theta$ from all previously evaluated configurations according to the best outcome of $f$ as the optimization result, we would be overly optimistic.
It is likely that a single best outcome can be partially attributed to the random error $\sigma^2$.
Instead, we are interested in the optimum of the true posterior mean.
Therefore, we employ the surrogate estimate of the posterior mean for each evaluated configuration to cancel out the noise.
The configuration for which the surrogate estimates the best outcome is then returned as the optimization result $\hat{\theta}^\ast$:
\begin{equation}
  \hat{\theta}^{\ast} := \operatorname{arg\,max}\limits_{\theta \in \mathcal{D}} \hat{\mu}(\theta),
\end{equation}
%To obtain a final outcome $\nu^\ast = f(\theta^\ast)$ that is independent of the optimization process, $f(\theta^\ast)$ is calculated again.
Using the stochastic outcome $\hat{\nu}^\ast := f(\hat{\theta}^\ast) $ observed during the optimization process would still potentially lead to an overly optimistic result.
Therefore an independent calculation of $f(\hat{\theta}^\ast)$ should be conducted to obtain a fair estimate of the true value of the best outcome.

@@

To compare \emph{BO} with a grid search that uses approximately the same time (see~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{BO} (see~\ref{tab:table_time}), with only a small advantage for \emph{Grid Small}.
Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{BO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{BO} and \emph{Grid}.
In the first case, the absolute differences are negligible as all methods lead to a high power.
For the second case we will later show, that the optimal configuration is close to the border of the search space.
The higher resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to the stochasticity.
These points are not included in the coarse grid.
In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.
The advantage of \emph{BO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect: paper2}, $n_{\text{treat}} = 500$.

@@

Figure~\ref{fig:plot_opt_path_5000} shows that the lower noise level leads to more stable simulation runs, meaning that each single optimization run leads to a results closer to the mean result.
Furthermore, the average optimization curve indicates that less iterations can be sufficient to obtain an outcome that is comparable to \emph{Grid},
It has to be noted, that this plot reports the best observed outcome so far and not the one that is chosen as the final result.
So on the left side of Figure~\ref{fig:plot_opt_path_5000}, we see more curves that observed overoptimistic outcomes in comparison the the right side.
Ideally, the choice of the final point should not be influenced by these overoptimistic observations for \emph{BO} as explained in Section~\ref{ssec:best_point}, as the final point is determined by the surrogates mean prediction.
This "bias correction" seems to be effective, as the validation error reported in Figure~\ref{fig:plot_boxplot_valid_y_5000} is nearly identical for \emph{BO} for the low and the high value of $n_\text{sim}$.
\begin{figure}[htb]
\centering
  \begin{minipage}{0.62\textwidth}
    \centering
    \includegraphics[width=\linewidth]{generated/figures/plot_opt_path_5000.pdf}%
    \captionof{figure}{Optimization progress for \emph{BO}: Best found outcome so far for each iteration in comparison to the best found outcome in the grid search on effect set \emph{paper} and $n_{\text{treat}} = 2000$.}
    \label{fig:plot_opt_path_5000}
  \end{minipage}\hspace{0.02\textwidth}%
  \begin{minipage}{0.35\textwidth}
    \centering
    \includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y_5000.pdf}%
    \captionof{figure}{Validated performance of all optimization strategies on effect set \emph{paper} and $n_{\text{treat}} = 2000$ with 100 stochastic repetitions and different values for \texttt{nsim}}
    \label{fig:plot_boxplot_valid_y_5000}
  \end{minipage}
\end{figure}
For all methods the IQR displayed in the Box plots is much smaller for $n_\text{sim} = 5000$, indicating that the found solutions vary less within the 100 stochastic replications.
Also the absolute differences between the methods seem even more negligible.
This demonstrates, that for the selected scenario all methods can find the optimal configuration if the results are not noisy.
But as an increased value of $n_\text{sim}$ also increases the runtime linearly, increasing $n_\text{sim}$ is not always a feasible choice if the goal is to find the optimal configuration quickly.
We conclude with the observation, that \emph{BO} can find configurations with a low number of needed simulations (higher noise) that are close to the optimal found solution with a high number of simulation.

@@

% subsection Summary
When planning clinical trials, determining patient group sizes and test methods is a crucial task.
The main goal is to maximize the test power, given a set of treatment and effect sizes under the null hypothesis.
If determining the test power involves intensive simulations, trying out different treatment design configurations can become a laborious task.
Either because large computational resources are required for an exhaustive exploration of all different configurations, or because it involves a tedious manual manual try and error process.
In this work we applied Bayesian Optimization to quickly find a clinical trial setup with high power from a large number of setup options.
We compared out approach to an exhaustive grid search by optimizing the power of adaptive seamless designs for different sets of treatment effect sizes.
BO was able to find competitive setups in a fraction of the time.
